---
title: "Hands-on Exercise 6A: Geographical Segmentation with Spatially Constrained Clustering Techniques"
# draft: false
date: "Sep 27, 2024"
date-modified: "last-modified"
author: Shubham Sinha
execute:
  echo: true
  eval: true
  freeze: true
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Overview

In this exercise, we will learn to delineate homogeneous regions using hierarchical and spatially constrained clustering techniques on geographically referenced multivariate data.

-   Convert GIS polygon data to R’s simple feature data frame.
-   Convert simple feature data frame to R’s SpatialPolygonDataFrame object.
-   Perform cluster analysis with *hclust()*.
-   Conduct spatially constrained cluster analysis using *skater()*.
-   Visualize analysis outputs using **ggplot2** and **tmap**.

## The Analytical Question

In geobusiness and spatial policy, delineating market or planning areas into homogeneous regions using multivariate data is a common approach.

> In this exercise, we aim to divide [Shan State, Myanmar](https://en.wikipedia.org/wiki/Shan_State) into homogeneous regions based on various Information and Communication Technology (ICT) indicators, such as radio, television, landline phones, mobile phones, computers, and home internet access.

## The Data

The following 2 datasets will be used in this study.

| **Data Set**                | **Description**                                                                                                         | **Format**     |
|----------------|----------------------------------------|----------------|
| myanmar_township_boundaries | GIS data in ESRI shapefile format containing township boundary information of Myanmar, represented as polygon features. | ESRI Shapefile |
| Shan-ICT.csv                | Extract of the 2014 Myanmar Population and Housing Census at the township level.                                        | CSV            |

Both datasets are downloaded from the [Myanmar Information Management Unit (MIMU)](http://themimu.info/).

## Installing and Launching the R Packages

The following R packages will be used in this exercise:

| **Package**   | **Purpose**                                                                                                | **Use Case in Exercise**                                                                                  |
|---------------|------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------|
| **sf**        | Handles vector-based geospatial data.                                                                       | Importing and manipulating township boundary data.                                                         |
| **spdep**     | Provides functions for spatial dependence analysis, including spatial weights and spatial autocorrelation.  | Performing spatially constrained cluster analysis using the SKATER method.                                  |
| **tmap**      | Creates static and interactive thematic maps using cartographic quality elements.                           | Visualizing geographic clusters and regional patterns in the township data.                                 |
| **tidyverse** | A collection of packages for data science tasks such as data manipulation, visualization, and modeling.     | Importing CSV files, wrangling data, performing data transformations, and visualizing analysis outputs.     |
| **corrplot**  | Provides visual tools to explore data correlation matrices.                                                 | Analyzing correlation between clustering variables.                                                        |
| **ggpubr**    | Enhances 'ggplot2' for easier 'publication-ready' plots.                                                    | Creating small multiple histograms and boxplots for exploratory data analysis.                              |
| **heatmaply** | Creates interactive cluster heatmaps.                                                                       | Visualizing multivariate data clustering results interactively.                                             |
| **cluster**   | Provides functions for cluster analysis.                                                                    | Performing hierarchical cluster analysis and evaluating clustering algorithms.                              |
| **ClustGeo**  | Implements Ward-like hierarchical clustering with spatial/geographical constraints.                        | Performing spatially constrained cluster analysis using a Ward-like hierarchical clustering algorithm.     |

To install and load these packages, use the following code:

```{r pacman}
pacman::p_load(spdep, tmap, sf, ClustGeo,
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally, plotly, knitr)
```


## Import Data and Preparation

### Import Geospatial Shapefile

Firstly, we will use [`st_read()`](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Myanmar Township Boundary shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r read_geospatial}
shan_sf <- st_read(dsn = "data/geospatial", layer = "myanmar_township_boundaries")
```

```{r view_shan_sf}
#| echo: false
kable(head(shan_sf))
```

Since our study area is the Shan state, we will examine the state list in the dataframe for the relevant state names (keys) for filtering.

```{r check_state_keys}
# Display unique values in the "ST" column sorted in ascending order
sort(unique(shan_sf$ST))
```

```{r filter_for_shan_state}
shan_sf <-
  shan_sf %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))

shan_sf
```

Then, we filter the dataframe to contain only the Shan states.

### Import Aspatial csv File

Next, we will import *Shan-ICT.csv* into R by using `read_csv()` of **readr** package. The output is R dataframe class.

```{r read_aspatial}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

```{r view_ict_stats}
summary(ict)
```

There are a total of 11 fields and 55 observation in the dataframe.

### Deriving New Variables Using the **dplyr** Package

The values in the dataset represent the number of households, which can be biased by the total number of households in each township. Townships with more households will naturally have higher numbers of households owning items like radios or TVs.

To address this bias, we will calculate the penetration rate for each ICT variable by dividing the number of households owning each item by the total number of households, then multiplying by 1000. This will normalize the data, allowing for a fair comparison between townships. Here is the code to perform this calculation:

```{r feat_eng}
ict_derived <- ict %>%
  # feature engineering: normalize data
  mutate(
    RADIO_PR = Radio / `Total households` * 1000,
    TV_PR = Television / `Total households` * 1000,
    LLPHONE_PR = `Land line phone` / `Total households` * 1000,
    MPHONE_PR = `Mobile phone` / `Total households` * 1000,
    COMPUTER_PR = Computer / `Total households` * 1000,
    INTERNET_PR = `Internet at home` / `Total households` * 1000
  ) %>%
  # improve readability of col names
  rename(
    DT_PCODE = `District Pcode`, DT = `District Name`,
    TS_PCODE = `Township Pcode`, TS = `Township Name`,
    TT_HOUSEHOLDS = `Total households`,
    RADIO = Radio, TV = Television,
    LLPHONE = `Land line phone`, MPHONE = `Mobile phone`,
    COMPUTER = Computer, INTERNET = `Internet at home`
  )
```

We can then review the summary statistics of the newly derived penetration rates using:

```{r view_feat_eng_stats}
summary(ict_derived)
```

This process adds six new fields to the data frame: `RADIO_PR`, `TV_PR`, `LLPHONE_PR`, `MPHONE_PR`, `COMPUTER_PR`, and `INTERNET_PR`, representing the penetration rates for each ICT variable.

## Exploratory Data Analysis (EDA)

### Statistical Graphics

Histogram is useful to identify the overall distribution of the data values (i.e. left skew, right skew or normal distribution)

```{r radio_ownership_hist}
p <- ggplot(data = ict_derived, aes(x = RADIO)) +
  geom_histogram(bins = 20, color = "black", fill = "lightblue") +
  labs(
    title = "Distribution of Radio Ownership",
    x = "Number of Households with Radio",
    y = "Frequency"
  ) +
  theme_minimal()

ggplotly(p)
```

Boxplot is useful to detect if there are outliers.

```{r radio_ownership_boxplot}
plot_ly(data = ict_derived,
        x = ~RADIO,
        type = 'box',
        fillcolor = 'lightblue',
        marker = list(color = 'lightblue'),
        notched = TRUE
       ) %>%
  layout(
    title = "Distribution of Radio Ownership",
    xaxis = list(title = "Number of Households with Radio")
  )
```

We will also plotting the distribution of the newly derived variables (i.e. Radio penetration rate).

```{r radio_pr_plot}
#| code-fold: true
#| code-summary: "Show the code"

mean_pr <- round(mean(ict_derived$`RADIO_PR`, na.rm = TRUE), 1)

# Define common axis properties
axis_settings <- list(
  title = "",
  zeroline = FALSE,
  showline = FALSE,
  showgrid = FALSE
)

aax_b <- list(
  title = "",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = FALSE,
  showgrid = FALSE
)

# Plot Histogram
histogram_plot <- plot_ly(
  ict_derived,
  x = ~`RADIO_PR`,
  type = 'histogram',
  histnorm = "count",
  marker = list(color = 'lightblue'),
  hovertemplate = 'Radio Penetration Rate: %{x}<br>Frequency: %{y}<extra></extra>'
) %>%
  add_lines(
    x = mean_pr, y = c(0, 15),
    line = list(width = 3, color = "black"),
    showlegend = FALSE
  ) %>%
  add_annotations(
    text = paste("Mean: ", mean_pr),
    x = mean_pr, y = 15,
    showarrow = FALSE,
    font = list(size = 12)
  ) %>%
  layout(
    xaxis = list(title = "Radio Penetration Rate"),
    yaxis = axis_settings,
    bargap = 0.1
  )

# Plot Boxplot
boxplot_plot <- plot_ly(
  ict_derived,
  x = ~`RADIO_PR`,
  type = "box",
  boxpoints = "all",
  jitter = 0.3,
  pointpos = -1.8,
  notched = TRUE,
  fillcolor = 'lightblue',
  marker = list(color = 'lightblue'),
  showlegend = FALSE
) %>%
  layout(
    xaxis = axis_settings,
    yaxis = aax_b
  )

# Combine Histogram and Boxplot
subplot(
  boxplot_plot, histogram_plot,
  nrows = 2, heights = c(0.2, 0.8), shareX = TRUE
) %>%
  layout(
    showlegend = FALSE,
    title = "Distribution of Radio Penetration Rate",
    xaxis = list(range = c(0, 500))
  )


```

::: callout-note
When we compare the distribution of radio ownership vs radio penetration rate, the distribution of radio penetration rate is less skewed.
:::

We can also plot multiple histograms to get a sense of the feature engineered fields.

```{r feat_eng_hist, fig.width=10}
#| code-fold: true
#| code-summary: "Show the code"

radio <- ggplot(data=ict_derived,
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")

tv <- ggplot(data=ict_derived,
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")

llphone <- ggplot(data=ict_derived,
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")

mphone <- ggplot(data=ict_derived,
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")

computer <- ggplot(data=ict_derived,
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")

internet <- ggplot(data=ict_derived,
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")

ggarrange(radio, tv, llphone, mphone, computer, internet, ncol = 3, nrow = 2)
```

### EDA using Choropleth Map

#### Joining Geospatial Data with Aspatial Data

Before we can prepare the choropleth map, we need to combine both the geospatial data object (i.e. *shan_sf*) and aspatial data.frame object (i.e. *ict_derived*) into one. This will be performed by using the [*left_join*](https://dplyr.tidyverse.org/reference/join.tbl_df.html) function of **dplyr** package. The *shan_sf* simple feature data.frame will be used as the base data object and the *ict_derived* data.frame will be used as the join table.

To join the data objects, we use the unique identifier, `TS_PCODE`.

```{r create_rds}
file_path <- "data/rds/shan_sf.rds"

if (!file.exists(file_path)) {
  shan_sf <- left_join(shan_sf, ict_derived, by = c("TS_PCODE" = "TS_PCODE"))
  write_rds(shan_sf, file_path)

} else {
  shan_sf <- read_rds(file_path)
}

kable(head(shan_sf))
```

#### Preparing a Choropleth Map

To have a quick look at the distribution of Radio penetration rate of Shan State at township level, a choropleth map will be prepared.

The code below is used to prepare the choropleth map by using the *qtm()* function of **tmap** package.

```{r radio_pr_choropleth}
qtm(shan_sf, "RADIO_PR")
```

In order to reveal the distribution shown in the choropleth map above are bias to the underlying total number of households at the townships, we will create two choropleth maps, one for the total number of households (i.e. TT_HOUSEHOLDS.map) and one for the total number of household with Radio (RADIO.map) by using the code below.

```{r households_radio_maps, fig.width=10}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) +
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks",
          title = "Total households") +
  tm_borders(alpha = 0.5)

RADIO.map <- tm_shape(shan_sf) +
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") +
  tm_borders(alpha = 0.5)

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)

```

Notice that the choropleth maps above clearly show that townships with relatively larger number ot households are also showing relatively higher number of radio ownership.

Now let us plot the choropleth maps showing the distribution of total number of households and Radio penetration rate by using the code below.

```{r plot_choropleth_households_radio, fig.width=10}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

::: callout-note
At first glance, for the Distribution of total number of households and radios plot, it reveals that township with relative larger number of households also exhibit relatively higher number of radio ownership.

The penetration rate plot provides more nuanced insight into radio ownership, highlighting regions where radios are more common among households, regardless of the total number of households in the region. This may suggest radio ownership rate is no sole dependent on the total population but also on socio-economic factors, or policies etc.
:::

