---
title: "Hands-on Exercise 8: Geographically Weighted Predictive Models"
author: "Christover Manafe"
date: "2024-10-16"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    number-depth: 4
---

# Overview

Predictive modeling uses statistical learning or machine learning techniques to forecast outcomes. Typically, the event being predicted occurs in the future. However, a set of known outcomes and predictors (also referred to as variables) is used to calibrate the predictive models.

This exercise aims to build predictive models using geographical random forest. Along the way, the intended acquired skills are:

-   Preparing training and test data sets using sampling methods
-   Calibrating predictive models using both geospatial statistical learning and machine learning methods
-   Comparing and selecting the best model for predicting the future outcome and,
-   Predicting future outcomes using the best calibrated model

# The Data

+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| **Dataset Name**                      | **Description**                                                                                                          | **Format**                                                                                                                              |
+=======================================+==========================================================================================================================+=========================================================================================================================================+
| Master Plan 2014 Subzone Boundary     | Geospatial data representing the boundaries of different areas in Singapore, specifically at the planning subzone level. | ESRI Shapefile                                                                                                                          |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=Master+Plan+2014+Subzone+Boundary+(Web)&page=1&resultId=d_5cb80a95445f236737f6bc2bfe5f159d) |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| HDB Resale data                       | Aspatial data containing HDB resale transacted prices in Singapore from Jan 2017 onwards.                                | CSV                                                                                                                                     |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=hdb+resale&page=1&resultId=d_8b84c4ee58e3cfc0ece0d773c8ca6abc)                              |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of eldercare in Singapore        | Geospatial data containing list of eldercare locations in Singapore.                                                     | geojson                                                                                                                                 |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=eldercare&page=1&resultId=d_3545b068e3f3506c56b2cb6b6117b884)                               |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of hawker centres in Singapore   | Geospatial data containing list of hawker centres locations in Singapore.                                                | geojson                                                                                                                                 |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=hawker+centre&page=1&resultId=d_4a086da0a5553be1d89383cd90d07ecd)                           |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of parks in Singapore            | Geospatial data containing list of parks locations in Singapore.                                                         | geojson                                                                                                                                 |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=parks&page=1&resultId=d_0542d48f0991541706b58059381a6eca)                                   |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of supermarkets in Singapore     | Geospatial data containing list of supermarkets locations in Singapore.                                                  | geojson                                                                                                                                 |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=supermarket&page=1&resultId=d_cac2c32f01960a3ad7202a99c27268a0)                             |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of CHAS clinics in Singapore     | Geospatial data containing list of CHAS clinics locations in Singapore.                                                  | geojson                                                                                                                                 |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=chas+clinics&page=1&resultId=d_548c33ea2d99e29ec63a7cc9edcccedc)                            |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of kindergartens in Singapore    | Geospatial data containing list of kindergartens locations in Singapore.                                                 | geojson                                                                                                                                 |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=childcare+services&page=1&resultId=d_5d668e3f544335f8028f546827b773b4)                      |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of MRT/LRT stations in Singapore | Geospatial data containing list of MRT/LRT stations locations in Singapore.                                              | ESRI Shapefile                                                                                                                          |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://datamall.lta.gov.sg/content/datamall/en/static-data.html)                                                              |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of bus stops in Singapore        | Geospatial data containing list of eldercare in Singapore.                                                               | ESRI Shapefile                                                                                                                          |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://datamall.lta.gov.sg/content/datamall/en/static-data.html)                                                              |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of primary school in Singapore   | Aspatial data containing list of primary schools in Singapore.                                                           | CSV                                                                                                                                     |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://data.gov.sg/datasets?query=primary+school&page=1&resultId=d_688b934f82c1059ed0a6993d2a829089)                          |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| CBD coordinates                       |                                                                                                                          | scraped from Google.                                                                                                                    |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| List of shopping malls in Singapore   | Aspatial data containing list of shopping malls in Singapore.                                                            | CSV                                                                                                                                     |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore)                                                             |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+
| Good primary schools                  | Aspatial data containing list of primary schools that are ordered in ranking in terms of popularity.                     | CSV                                                                                                                                     |
|                                       |                                                                                                                          |                                                                                                                                         |
|                                       |                                                                                                                          | [Source](https://www.salary.sg/2021/best-primary-schools-2021-by-popularity)                                                            |
+---------------------------------------+--------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------+

: {tbl-colwidths="\[25,50,25\]"}

# The Packages

We will use following packages in this exercise

::: panel-tabset
## Packages

We will use following packages in this exercise:

+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Package**                                                               | **Description**                                                                                                                                                                                                   |
+===========================================================================+===================================================================================================================================================================================================================+
| [**sf**](https://r-spatial.github.io/sf/)                                 | Provides functions to manage, process, and manipulate **Simple Features**, a formal geospatial data standard that specifies a storage and access model of spatial geometries such as points, lines, and polygons. |
+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**spdep**](https://cran.r-project.org/web/packages/spdep/)               | Provides a collection of functions to create spatial weights matrix objects from polygon 'contiguities', point patterns by distance, and tessellations.                                                           |
+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**tidyverse**](https://www.tidyverse.org/)                               | A collection of R packages for data science tasks such as importing, tidying, wrangling, and visualizing data.                                                                                                    |
+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**tmap**](https://cran.r-project.org/web/packages/tmap/)                 | Provides functions for creating cartographic-quality static maps or interactive maps using the [leaflet](https://leafletjs.com/) API.                                                                             |
+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/index.html) | A package for calibrating the geographically weighted family of models.                                                                                                                                           |
+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**SpatialML**](https://cran.r-project.org/web/packages/SpatialML/)       | A package that implements a spatial extension of the random forest algorithm.                                                                                                                                     |
+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**rsample**](https://rsample.tidymodels.org/)                            | A package that provides functions to create different types of resamples and corresponding classes for their analysis.                                                                                            |
+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**Metrics**](https://github.com/mfrasco/Metrics)                         | A packages for common machine learning metrics.                                                                                                                                                                   |
+---------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: {tbl-colwidths="\[15,85\]"}

## Code

To install and launch all R packages.

```{r}
#|eval: false
pacman::p_load(sf, spdep, GWmodel, SpatialML, tmap, rsample, Metrics, tidyverse)
```
:::

# Data Preparation

## Import the data

We already have processed data on above datasets in .rds format. Let's load it using `read_rds()` function.

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
glimpse(mdata)
```

## Data Sampling

The entire dataset is split into training and test sets, with 65% allocated to the training set and 35% to the test set, using the `initial_split()` function from **rsample** package.

```{r}
set.seed(1234)
resale_split <- initial_split(mdata, 
                              prop = 6.5/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

```{r}
write_rds(train_data, "data/model/train_data.rds")
write_rds(test_data, "data/model/test_data.rds")
```

# Compute Correlation Matrix

Before loading the predictors into a predictive model, it is good practice to examine a correlation matrix to check for signs of multicollinearity.

```{r fig.height=8}
mdata_nogeo <- mdata %>%
  st_drop_geometry()
corrplot::corrplot(cor(mdata_nogeo[, 2:17]), 
                   diag = FALSE, 
                   order = "AOE",
                   tl.pos = "td", 
                   tl.cex = 0.5, 
                   method = "number", 
                   type = "upper")
```

> The correlation matrix above shows that all correlation values are below 0.8, indicating no signs of multicollinearity.

# Non-Spatial Multiple Linear Regression

In this section, we will build a non-spatial multiple linear regression model. This statistical technique uses several explanatory variables to predict the outcome of a response variable. The goal is to model the relationship between the explanatory and response variables.

We will use the `lm()` function to construct the model.

```{r}
price_mlr <- lm(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
summary(price_mlr)
```

# GWR Predictive Method

In this section, we will learn how to calibrate a model to predict HDB resale prices using the geographically weighted regression (GWR) method from the [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/index.html) package.

## Convert sf data frame to SpatialPointDataFrame

First, we need to convert our `sf` data frame to a `SpatialPointsDataFrame`, as the functions in the `GWmodel` package require data in this format. We can perform this conversion using the `as_Spatial()` function from the `sf` package.

```{r}
train_data_sp <- as_Spatial(train_data)
train_data_sp
```

## Compute adaptive bandwidth

Next, the `bw.gwr()` function from the `GWmodel` package will be used to determine the optimal bandwidth for the GWR model.

```{r}
#| eval: false
bw_adaptive <- bw.gwr(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                  data=train_data_sp,
                  approach="CV",
                  kernel="gaussian",
                  adaptive=TRUE,
                  longlat=FALSE)
write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

> In the adaptive approach, the bandwidth is not a fixed distance but is determined by the number of nearest neighbor points. This approach is particularly useful in areas where the density of data points varies.
>
> The result from the `bw.gwr()` function indicates that the optimal bandwidth for this dataset is 40 neighbor points. This means that when estimating the parameters for a given location, the model will consider the 40 nearest neighbors.

## Construct adaptive bandwidth GWR model

Next, let us call the saved bandwidth by using the code chunk below.

```{r}
bw_adaptive <- read_rds("data/model/bw_adaptive.rds")
bw_adaptive
```

Now, we can proceed to calibrate the GWR-based hedonic pricing model using an adaptive bandwidth and a Gaussian kernel, as shown in the code chunk below.

```{r}
#| eval: false
gwr_adaptive <- gwr.basic(formula = resale_price ~
                            floor_area_sqm + storey_order +
                            remaining_lease_mths + PROX_CBD + 
                            PROX_ELDERLYCARE + PROX_HAWKER +
                            PROX_MRT + PROX_PARK + PROX_MALL + 
                            PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                            WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                            WITHIN_1KM_PRISCH,
                          data=train_data_sp,
                          bw=bw_adaptive, 
                          kernel = 'gaussian', 
                          adaptive=TRUE,
                          longlat = FALSE)
write_rds(gwr_adaptive, "data/model/gwr_adaptive.rds")
```

## Retrieve GWR output object

The code chunk below will be used to retrieve the saved GWR model object.

```{r}
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
gwr_adaptive
```

## **Convert test data from sf data frame to SpatialPointDataFrame**

We need to convert our test data from an `sf` data frame to a `SpatialPointsDataFrame`, as the functions in the `GWmodel` package require data in this format. We will use the `as_Spatial()` function from the **sf** package to perform this conversion.

```{r}
test_data_sp <- test_data %>%
  as_Spatial()

test_data_sp
```

## **Compute adaptive bandwidth for test data**

Next, similar to the process used for the training data, we will use the `bw.gwr()` function from the **GWmodel** package to determine the optimal bandwidth for the GWR model on the test data.

```{r}
#| eval: false
gwr_bw_test_adaptive <- bw.gwr(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                  data=test_data_sp,
                  approach="CV",
                  kernel="gaussian",
                  adaptive=TRUE,
                  longlat=FALSE)
write_rds(gwr_bw_test_adaptive, "data/model/gwr_bw_test_adaptive.rds")
```

```{r}
#| echo: false
gwr_bw_test_adaptive <- read_rds("data/model/gwr_bw_test_adaptive.rds")
```

## Compute predicted values of test data

Finally, we use the `gwr.predict()` function from the **`GWmodel`** package to compute the predicted values for the test data based on our GWR model.

```{r}
#| eval: false
gwr_pred <- gwr.predict(formula = resale_price ~
                          floor_area_sqm + storey_order +
                          remaining_lease_mths + PROX_CBD + 
                          PROX_ELDERLYCARE + PROX_HAWKER + 
                          PROX_MRT + PROX_PARK + PROX_MALL + 
                          PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                          WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
                          WITHIN_1KM_PRISCH, 
                        data=train_data_sp, 
                        predictdata = test_data_sp, 
                        bw=40, 
                        kernel = 'gaussian', 
                        adaptive=TRUE, 
                        longlat = FALSE)
```

# Preparing coordinates data

## Extracting coordinates data

The code chunk below extract the x, y coordinates of the full, training and test data sets using `st_coordinates()` function from the **sf** package.

```{r}
coords <- st_coordinates(mdata)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)
```

Save the output into rds for future use.

```{r}
#| eval: false
coords_train <- write_rds(coords_train, "data/model/coords_train.rds" )
coords_test <- write_rds(coords_test, "data/model/coords_test.rds" )
```

```{r}
#| echo: false
coords_train <- read_rds("data/model/coords_train.rds")
coords_test <- read_rds("data/model/coords_test.rds")
```

## Dropping geometry field

Next, we drop the geometry column from the `sf` data frame, as it is not needed for our upcoming analysis. We will use the `st_drop_geometry()` function from the `sf` package to perform this operation.

```{r}
train_data <- train_data %>% 
  st_drop_geometry()
```

# Calibrating Random Forest Model

In this section, we will learn how to calibrate a model to predict HDB resale prices using the random forest function from the [**ranger**](https://cran.r-project.org/web/packages/ranger/index.html) package. After fitting the model, we save it as an RDS file using the `write_rds()` function.

```{r}
#| eval: false
set.seed(1234)
rf <- ranger(resale_price ~ floor_area_sqm + storey_order + 
               remaining_lease_mths + PROX_CBD + PROX_ELDERLYCARE + 
               PROX_HAWKER + PROX_MRT + PROX_PARK + PROX_MALL + 
               PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
               WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
               WITHIN_1KM_PRISCH,
             data=train_data)
write_rds(rf, "data/model/rf.rds")
```

```{r}
#| echo: false
rf <- read_rds("data/model/rf.rds")
```

```{r}
rf
```

# Calibrating Geographical Random Forest Model

In this section, we will learn how to calibrate a model to predict HDB resale price using the `grf()` function of [**SpatialML**](https://cran.r-project.org/web/packages/ranger/index.html) package.

## Calibrating using training data

Let's calibrate a geographic random forest model then save the model.

```{r}
#| eval: false
set.seed(1234)
gwRF_adaptive <- grf(formula = resale_price ~ floor_area_sqm + storey_order +
                       remaining_lease_mths + PROX_CBD + PROX_ELDERLYCARE +
                       PROX_HAWKER + PROX_MRT + PROX_PARK + PROX_MALL +
                       PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                       WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                       WITHIN_1KM_PRISCH,
                     dframe=train_data, 
                     bw=55,
                     kernel="adaptive",
                     coords=coords_train)
write_rds(gwRF_adaptive, "data/model/gwRF_adaptive.rds")
```

```{r}
#| echo: false
gwRF_adaptive <- read_rds("data/model/gwRF_adaptive.rds")
```

## Predicting using test data

### Preparing test data

First, combine the test data with the corresponding coordinates. Since none of the rows in our datasets were sorted, using `cbind()` is sufficient. The `st_drop_geometry()` function is used to remove the geometry column.

```{r}
test_data <- cbind(test_data, coords_test) %>%
  st_drop_geometry()
```

### Predicting with test data

Next, the `predict.grf()` function from the **SpatialML** package will be used to predict the resale value using the test data and the previously calibrated `gwRF_adaptive` model.

```{r}
#| eval: false
GRF_pred <- predict.grf(gwRF_adaptive, 
                           test_data, 
                           x.var.name="X",
                           y.var.name="Y", 
                           local.w=1,
                           global.w=0)
write_rds(GRF_pred, "data/model/GRF_pred.rds")
```

### Converting prediction output into a data frame

The output of `predict.grf()` is a vector of predicted values. It is advisable to convert this into a data frame for easier visualization and analysis.

```{r}
GRF_pred <- read_rds("data/model/GRF_pred.rds")
GRF_pred_df <- as.data.frame(GRF_pred)
```

Then, we'll use `cbind()` to append the predicted values to the test data.

```{r}
test_data_p <- cbind(test_data, GRF_pred_df)
write_rds(test_data_p, "data/model/test_data_p.rds")
```

## Calculate Root Mean Square Error (RMSE)

The root mean square error (RMSE) allows us to measure how far the predicted values are from the observed values in a regression analysis. In the code chunk below, the `rmse()` function from the `Metrics` package is used to compute the RMSE.

```{r}
rmse(test_data_p$resale_price,
     test_data_p$GRF_pred)
```

## Visualize Predicted Value

A scatterplot can be used to visualize the actual resale prices versus the predicted resale prices, as shown in the code chunk below.

```{r}
ggplot(data = test_data_p,
       aes(x = GRF_pred,
           y = resale_price)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "solid")
```

# Reference

Kam, T. S. Geographically Weighted Predictive Models. *R for Geospatial Data Science and Analytics.* <https://r4gdsa.netlify.app/chap14.html>