---
title: "7A: Calibrating Hedonic Pricing Model for Private Highrise Property with GWR Method"
# draft: true
date: "October 10, 2024"
date-modified: "last-modified"
author: Shubham Sinha
execute:
  echo: true
  eval: true
  freeze: true
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Overview

In this exercise, we will learn to build [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp) models for private high-rise property using Geographically Weighted Regression (GWR) methods to account for non-stationary variables.

**Geographically weighted regression (GWR)** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable).

> In this exercise, The dependent variable is the resale prices of condominium in 2015. The independent variables are divided into either structural and locational.

We will be using the [**GWmodel**](https://www.jstatsoft.org/article/view/v063i17) package. It provides a collection of localised spatial statistical methods, namely: GW summary statistics, GW principal components analysis, GW discriminant analysis and various forms of GW regression; some of which are provided in basic and robust (outlier resistant) forms. Commonly, outputs or parameters of the GWmodel are mapped to provide a useful exploratory tool, which can often precede (and direct) a more traditional or sophisticated statistical analysis.

## Learning Outcome

-   Understand the basics of geographically weighted regression (GWR).
-   Build and evaluate hedonic pricing models using GWR.
-   Convert geospatial and aspatial data into appropriate formats for analysis.
-   Perform exploratory data analysis (EDA) using statistical graphics.
-   Visualize the model outputs and spatial patterns using various R packages.
-   Assess spatial autocorrelation and other statistical properties of model residuals.
-   Compare and interpret fixed and adaptive bandwidth GWR models.

## The Data

The following 2 datasets will be used in this study:

| **Data Set** | **Description** | **Format** |
|----------------|-----------------------------------------|----------------|
| MP14_SUBZONE_WEB_PL | URA Master Plan 2014's planning subzone boundaries represented as polygon features. | ESRI Shapefile |
| condo_resale_2015.csv | Data on condominium resale prices in 2015, including structural and locational attributes. | CSV |

## Installing and Launching the R Packages

The following R packages will be used in this exercise:

| **Package** | **Purpose** | **Use Case in Exercise** |
|-----------------|-----------------------------|--------------------------|
| **sf** | Manages, processes, and manipulates vector-based geospatial data. | Handling and converting geospatial data such as the URA Master Plan boundaries. |
| **GWmodel** | Calibrates geographically weighted models for local spatial analysis. | Building hedonic pricing models using fixed and adaptive bandwidth GWR methods. |
| **olsrr** | Performs diagnostics and builds better multiple linear regression models. | Testing for multicollinearity, normality, and linearity of the regression model. |
| **corrplot** | Provides visual tools to explore data correlation matrices. | Visualizing relationships between independent variables in the dataset. |
| **tmap** | Creates static and interactive thematic maps using cartographic quality elements. | Visualizing the geospatial distribution of condominium resale prices and other model outputs. |
| **tidyverse** | A collection of packages for data science tasks such as data manipulation, visualization, and modeling. | Importing CSV files, wrangling data, and performing data transformations and visualizations. |
| **ggpubr** | Enhances 'ggplot2' for easier 'publication-ready' plots. | Creating small multiple histograms and scatterplots for exploratory data analysis. |

To install and load these packages, use the following code:

```{r pacman}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary)
```

## Geospatial Data Wrangling

### Importing Geospatial Data

To import MP_SUBZONE_WEB_PL shapefile:

```{r}
mpsz = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

The output above shows that the R object used to contain the imported MP14_SUBZONE_WEB_PL shapefile is called *mpsz* and it is a simple feature object. The geometry type is *multipolygon*. It is also important to note that **mpsz simple feature object does not have EPSG information**.

### Updating CRS Information

The code below updates the newly imported *mpsz* with the correct ESPG code (i.e. 3414)

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
st_crs(mpsz_svy21)
```

Notice that the EPSG: is indicated as *3414* now.

### Revealing the Extent of mpsz_svy21

To reveal the extent of *mpsz_svy21*:

```{r}
st_bbox(mpsz_svy21) #view extent
```

## Aspatial Data Wrangling

### Importing Aspatial Data

The *condo_resale_2015* is in csv file format. The codes chunk below uses `read_csv()` function of **readr** package to import *condo_resale_2015* into R as a tibble data frame called *condo_resale*.

```{r}
condo_resale = read_csv("data/aspatial/Condo_resale_2015.csv")
glimpse(condo_resale)
```

To display summary statistics of *condo_resale*:

```{r}
summary(condo_resale)
```

### Converting an Aspatial Data Frame to an sf Object

In this step, we will:

-   Use the `st_as_sf()` function from the **sf** package to convert the aspatial data frame into a spatial (sf) object.
-   Apply `st_transform()` to reproject the coordinates from the WGS 84 coordinate system (CRS: 4326) to SVY21 (CRS: 3414), commonly used in Singapore.

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs = 4326) %>%
  st_transform(crs = 3414)

head(condo_resale.sf)
```

-   The resulting output is a point-based feature data frame.

## Exploratory Data Analysis

In the section, we will use statistical graphics functions of `ggplot2` package to perform EDA.

### EDA using Statistical Graphics

### 6.1.1 Plot Distribution

To plot the distribution of *SELLING_PRICE* by using histograms:

```{r selling_price_hist, fig.width=12, fig.height=8}
ggplot(data=condo_resale.sf, aes(x=`SELLING_PRICE`)) +
  geom_histogram(bins=20, color="black", fill="light blue")
```

::: callout-note
**Observations:**

-   A right skewed distribution.
-   This means that more condominium units were transacted at relative lower prices.
-   Statistically, the skewed distribution **can be normalised by using log transformation**.
:::

```{r log_selling_price_hist, fig.width=12, fig.height=8}
condo_resale.sf <- condo_resale.sf %>%
  mutate(`LOG_SELLING_PRICE` = log(SELLING_PRICE))

ggplot(data=condo_resale.sf, aes(x=`LOG_SELLING_PRICE`)) +
  geom_histogram(bins=20, color="black", fill="light blue")
```

Notice that the distribution is relatively **less skewed after the transformation.**

### Multiple Histogram Plots for Variable Distribution

In this section, we will create multiple histograms (also known as a trellis plot or small multiples) to visualize the distribution of several variables. We will use the `ggarrange()` function from the [**ggpubr**](https://cran.r-project.org/web/packages/ggpubr/index.html) package to organize these histograms into a 3-column by 4-row grid layout.

```{r, trellis_condo_resale, fig.width=12, fig.height=8}
#| code-fold: true
#| code-summary: "Show the code"

# List of variables to plot
variables <- c("AREA_SQM", "AGE", "PROX_CBD", "PROX_CHILDCARE", 
               "PROX_ELDERLYCARE", "PROX_URA_GROWTH_AREA", 
               "PROX_HAWKER_MARKET", "PROX_KINDERGARTEN", 
               "PROX_MRT", "PROX_PARK", "PROX_PRIMARY_SCH", 
               "PROX_TOP_PRIMARY_SCH")

histograms <- lapply(variables, function(var) {
  ggplot(condo_resale.sf, aes_string(x = var)) + 
    geom_histogram(bins = 20, color = "black", fill = "lightblue")
})

ggarrange(plotlist = histograms, ncol = 3, nrow = 4)
```
