---
title: "Hands-on Exercise 3 - Spatio-Temporal Point Patterns Analysis"
author: "Shubham Sinha"
date: "2024-09-14"
date-modified: "last-modified"
---

## 1.Overview

Spatio-temporal point processes model the occurrence of events across both space and time, capturing phenomena such as disease outbreaks, natural disasters, or species sightings. With the growing availability of geographically and temporally tagged data, analyzing these patterns has become essential across disciplines. This chapter introduces a practical approach to spatio-temporal point pattern analysis using R, showcasing how multiple packages can be integrated to perform robust and intuitive analyses. To illustrate the methodology, it applies these tools to forest fire events recorded in Kepulauan Bangka Belitung, Indonesia, throughout 2023, highlighting key techniques, workflows, and interpretations.

## 2. Data

-   Analyze spatio-temporal patterns of forest fires in Kepulauan Bangka Belitung, Indonesia.

-   Both datasets are filtered to include only data within Kepulauan Bangka Belitung.

-   Enables spatial overlay and temporal analysis of fire activity within administrative boundaries.

-   Use two datasets:

    -   forestfires.csv: Contains geolocated fire events detected by the MODIS satellite sensor.

    -   Kepulauan_Bangka_Belitung shapefile: Defines sub-district (kelurahan) boundaries of the region.
    
## 3. Installing and Loading packages

```{r}
pacman::p_load(sf, terra, spatstat, sparr, tmap, tidyverse)
```

## 4. Importing & Preparing Study Area

### 4.1 Importing study area

```{r}
kbb <- st_read(dsn="data",
               layer = "Kepulauan_Bangka_Belitung") 
```

Cleaning & Preparing the geometry

```{r}
kbb_sf <- st_read(dsn="data",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

Inspect the Result

```{r}
glimpse(kbb_sf)
plot(kbb_sf)
```

### 4.2 Converting OWIN

Confirming sf to Owin and inspecting 

```{r}
kbb_owin <- as.owin(kbb_sf)
```
```{r}
kbb_owin
```

Confirm the class 

```{r}
class(kbb_owin)
```

## 5. Importing the Preparing Forest Fire Data

Import CSV and Convert to Spatial Format

Use read_csv() to load the fire data and convert it to an sf object using longitude and latitude.

```{r}
fire_sf <- read_csv("data/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 32748)  # Reproject to UTM Zone 48S
```

This creates a spatial object with fire points in the correct coordinate system for analysis.

Prepare Temporal Attributes

Since ppp objects only accept numeric or character marks, you’ll extract and format the date info:

```{r}
fire_sf <- fire_sf %>%
  mutate(
    DayofYear = yday(acq_date),              # Numeric day of year
    Month_num = month(acq_date),             # Numeric month
    Month_fac = month(acq_date,              # Factor month (e.g., "January")
                      label = TRUE, 
                      abbr = FALSE)
  )
```


## 6. Visualising the Fire Points

### 6.1 Overall Plot

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

### 6.2 Visuaising geographic distribution of forest fires by month

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

## 7. Computing STKDE by Month
Computing STKDE by Month with the same clarity and structure. This step sets the stage for spatio-temporal kernel density estimation (STKDE) using the {sparr} package, which is powerful for visualizing fire intensity across space and time.

### 7.1 Extracting forest fire by month

```{r}
fire_month <- fire_sf %>%
  select(Month_num)
```

### 7.2 Creating ppp

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

```{r}
summary(fire_month_ppp)
```

```{r}
any(duplicated(fire_month_ppp))
```

### 7.3 Including the OWIN Object

Combine ppp Object with owin Window

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
```

Inspect the Resulting Object

```{r}
summary(fire_month_owin)
```

Visualize the Spatial Window

```{r}
plot(fire_month_owin)
```

This gives you a sanity check before diving into density estimation.

### 7.4 Computing Spatio-temporal KDE

spattemp.density() of sparr package is used to compute the STKDE.

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

### 7.5 Plotting the spatio-temporal KDE object

plot() of R base is used to the KDE for between July 2023 - December 2023.

```{r}
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

## 8. Computing STKDE by Day of Year

### 8.1 Creating ppp object

 DayofYear field is included in the output ppp object.

```{r}
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

### 8.2 Including Owin object
Combining the ppp object and the owin object.

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

### 8.3 Visualizations of the plots

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```

```{r}
plot(kde_yday)
```

## 9. Computing STKDE by Day of Year: Improved method

One of the nice function provides in sparr package is BOOT.spattemp(). It support bandwidth selection for standalone spatiotemporal density/intensity based on bootstrap estimation of the MISE, providing an isotropic scalar spatial bandwidth and a scalar temporal bandwidth.

Code chunk below uses BOOT.spattemp() to determine both the spatial bandwidth and the scalar temporal bandwidth.

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin) 
```

### 9.1 Computing spatio-temporal KDE
The STKDE will be derived by using h and lambda values derive in previous step.

```{r}
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 9000,
                             lambda = 19)
summary(kde_yday)
```

### 9.2 Plotting the output spatio temporal KDE

```{r}
plot(kde_yday)
```