---
title: "Hands-on Exercise 5B: Local Measures of Spatial Autocorrelation"
author: "Shubham Sinha"
date: "2025-09-23"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    number-depth: 4
---

## Overview

Local Measures of Spatial Autocorrelation (LMSA) focus on the relationships between each observation and its surroundings, rather than providing a single summary of these relationships across the map. In this sense, they are not summary statistics but scores that allow us to learn more about the spatial structure in our data. The general intuition behind the metrics however is similar to that of global ones. Some of them are even mathematically connected, where the global version can be decomposed into a collection of local ones. One such example are Local Indicators of Spatial Association (LISA). Beside LISA, Getis-Ord’s Gi-statistics will be introduce as an alternative LMSA statistics that present complementary information or allow us to obtain similar insights for geographically referenced data.

In this hands-on exercise, you will learn how to compute Local Measures of Spatial Autocorrelation (LMSA) by using sfdep package. By the end to this hands-on exercise, you will be able to:

- Import geospatial data using appropriate function(s) of sf package,

- Import csv file using appropriate function of readr package,

- Perform relational join using appropriate join function of dplyr package,

- Compute Local Indicator of Spatial Association (LISA) statistics for detecting clusters and outliers by using appropriate functions sfdep package;

- Compute Getis-Ord’s Gi-statistics for detecting hot spot or/and cold spot area by using appropriate functions of sfdep package; and

- Visualise the analysis output by using tmap package.

## The Analytical Question

In spatial policy planning, one of the main development objective of the local government and planners is to ensure equal distribution of development in the province. In this study, we will apply spatial statistical methods to examine the distribution of development in Hunan Province, China, using a selected indicator (e.g., GDP per capita).

> Our key questions are:
>
> 1.  Is development evenly distributed geographically?
> 2.  If not, is there evidence of spatial clustering?
> 3.  If clustering exists, where are these clusters located?

## The Data

The following 2 datasets will be used in this exercise.

| **Data Set**                | **Description**                                                  | **Format**     |
|------------------|------------------------------------|------------------|
| Hunan county boundary layer | Geospatial data set representing the county boundaries of Hunan  | ESRI Shapefile |
| Hunan_2012.csv              | Contains selected local development indicators for Hunan in 2012 | CSV            |

## Installing and Launching the R Packages

The following R packages will be used in this exercise:

| **Package**           | **Purpose**                                                                                     | **Use Case in Exercise**                                                                   |
|-----------------------|-------------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------------|
| **sf**                | Handles spatial data, particularly vector-based geospatial data.                                | Importing and managing township boundary data for Myanmar.                                  |
| **rgdal**             | Provides bindings to the Geospatial Data Abstraction Library (GDAL) for reading and writing spatial data. | Reading and writing geospatial data in various formats, including shapefiles.               |
| **spdep**             | Analyzes spatial dependence and provides tools for spatial econometrics.                        | Performing spatially constrained clustering and other spatial dependence analyses.          |
| **tidyverse**         | A collection of packages for data science tasks like data manipulation and visualization.       | Handling attribute data, reading CSV files, and data wrangling with **readr**, **dplyr**, and **ggplot2**. |
| **tmap**              | Creates static and interactive thematic maps.                                                   | Visualizing data using choropleth maps to display spatial patterns and relationships.       |
| **corrplot**          | Visualizes correlation matrices.                                                                | Creating correlation plots to explore relationships between different ICT measures.         |
| **ggpubr**            | Provides functions to create and customize 'ggplot2'-based publication-ready plots.              | Enhancing multivariate data visualizations for clearer presentation of results.             |
| **heatmaply**         | Generates interactive heatmaps.                                                                 | Visualizing multivariate data through interactive heatmaps for deeper insights.             |
| **cluster**           | Performs cluster analysis.                                                                      | Conducting hierarchical clustering to group similar regions based on ICT measures.          |
| **ClustGeo**          | Performs spatially constrained hierarchical clustering.                                         | Applying spatial constraints to hierarchical clustering for identifying homogeneous regions. |
To install and load these packages, use the following code:

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```


## Import Data and Preparation

### Import Geospatial Shapefile

Firstly, we will use [`st_read()`](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

### Import Aspatial csv File

Next, we will import *Hunan_2012.csv* into R by using `read_csv()` of **readr** package. The output is R dataframe class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### Perform Relational Join

Then, we will perform a `left_join()` to update the attribute table of hunan’s SpatialPolygonsDataFrame with the attribute fields of hunan2012 dataframe.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

### Visualizing Regional Development Indicator

Now to build two choropleth maps by using equal interval (i.e equal) and quantile (i.e. quantile) classification methods, and
to plot both maps next to each other by using tmap_arrange().

```{r}
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile interval classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## Local Indicators of Spatial Association(LISA)

Local Indicators of Spatial Association (LISA) are statistics used to identify clusters and outliers in the spatial distribution of a variable. For example, if we are analyzing the GDP per capita in Hunan Province, China, LISA can help detect areas (counties) where GDP values are significantly higher or lower than expected by chance. This means that these values deviate from what would be seen in a random distribution across space.

In this section, we will apply appropriate Local Indicators for Spatial Association (LISA), particularly the Local Moran's I statistic, to identify clusters and outliers in the 2012 GDP per capita data for Hunan Province.

### Computing Contiguity Spatial Weights

Before we can compute the global spatial autocorrelation statistics, we need to construct a spatial weights of the study area. The spatial weights is used to define the neighbourhood relationships between the geographical units (i.e. county) in the study area.

In spatial data analysis using the `sfdep` package, the process of constructing spatial weight matrices begins with the `st_contiguity()` function. This function identifies neighboring polygons based on shared boundaries and allows the user to specify the type of contiguity—Queen or Rook—through the `queen` argument. By default, it uses Queen contiguity, meaning any shared point or edge qualifies two regions as neighbors unless explicitly set to `queen = FALSE`. Once the neighbor list is created, the `st_weights()` function is used to compute spatial weights, which quantify the influence of neighboring regions in spatial modeling.

---

- **`st_contiguity()` Function:**
  - Computes contiguity-based neighbor lists.
  - Uses shared boundaries to define neighbors.
  - `queen = TRUE` (default): Queen contiguity (shared point or edge).
  - `queen = FALSE`: Rook contiguity (shared edge only).

- **`st_weights()` Function:**
  - Converts neighbor list (`nb`) into spatial weights.
  - **Arguments:**
    - `nb`: Neighbor list object from `st_neighbors()`.
    - `style`: Defines weight calculation method:
      - `"W"`: Row-standardized (default).
      - `"B"`: Binary coding.
      - `"C"`: Global standardization.
      - `"U"`: Global standardization divided by number of neighbors.
      - `"minmax"`: Min-max normalization.
      - `"S"`: Variance-stabilizing (Tiefelsdorf et al., 1999).
    - `allow_zero`: If `TRUE`, assigns zero to zones with no neighbors.

```{r}
library(sfdep)
wm_q <- hunan %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(
           nb, style = "W"),
         .before = 1)
```

```{r}
summary(wm_q$nb)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbours.

### Row-standardised Weights Matrix

To assign spatial weights to neighboring polygons, the style = "W" method is commonly used, which distributes equal weight by assigning each neighbor a value of \frac{1}{\text{number of neighbors}}. This approach is intuitive and straightforward for summarizing neighboring values like income, but it can introduce bias for edge polygons that have fewer neighbors, potentially distorting spatial autocorrelation. Despite this limitation, style = "W" is used here for simplicity, though more robust alternatives like style = "B" exist.


::: callout-tip
For simplicity, we use the style="W" option in this example, but keep in mind that other, potentially more accurate methods are available, such as style="B".
:::


::: callout-tip
-   The input of [`nb2listw()`](https://r-spatial.github.io/spdep/reference/nb2listw.html) must be an object of class **nb**. The syntax of the function has two major arguments, namely style and zero.poly.

-   *style* can take values "W", "B", "C", "U", "minmax" and "S". B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   The zero.policy=TRUE option allows for lists of non-neighbors. This should be used with caution since the user may not be aware of missing neighbors in their dataset however, a zero.policy of FALSE would return an error.
:::

### Computing Local Moran's I

To compute local Moran's I, the [*localmoran()*](https://r-spatial.github.io/spdep/reference/localmoran.html) function of **spdep** will be used. It computes *Ii* values, given a set of *zi* values and a listw object providing neighbour weighting information for the polygon associated with the zi values.


Now, we compute local Moran's I of *GDPPC2012* at the county level.

```{r}
lisa <- wm_q %>% 
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

::: callout-tip
*localmoran()* function returns a matrix of values whose columns are:

-   Ii: the local Moran's I statistics
-   E.Ii: the expectation of local moran statistic under the randomisation hypothesis
-   Var.Ii: the variance of local moran statistic under the randomisation hypothesis
-   Z.Ii:the standard deviate of local moran statistic
-   Pr(): the p-value of local moran statistic
:::

Next, we list the content of the local Moran matrix derived by using [*printCoefmat()*](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/printCoefmat).

```{r}
glimpse(lisa)
```

#### Mapping the Local Moran's I
