---
title: "Hands-on Exercise 5A: Global Measures of Spatial Autocorrelation"
author: "Shubham Sinha"
date: "2025-09-21"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    number-depth: 4
---

## Overview

In this hands-on exercise, you will learn how to compute Global Measures of Spatial Autocorrelation (GMSA) by using **spdep** package. By the end to this hands-on exercise, you will be able to:

-   Import geospatial data using appropriate function(s) of **sf** package,
-   Import csv file using appropriate function of **readr** package,
-   Perform relational join using appropriate join function of **dplyr** package,
-   Compute Global Spatial Autocorrelation (GSA) statistics by using appropriate functions of **spdep** package,
    -   plot Moran scatterplot,
    -   compute and plot spatial correlogram using appropriate function of **spdep** package.
-   provide statistically correct interpretation of GSA statistics.

## The Analytical Question

In spatial policy planning, one of the main development objective of the local government and planners is to ensure equal distribution of development in the province. In this study, we will apply spatial statistical methods to examine the distribution of development in Hunan Province, China, using a selected indicator (e.g., GDP per capita).

> Our key questions are:
>
> 1.  Is development evenly distributed geographically?
> 2.  If not, is there evidence of spatial clustering?
> 3.  If clustering exists, where are these clusters located?

## The Data

The following 2 datasets will be used in this exercise.

| **Data Set** | **Description** | **Format** |
|-----------------|-------------------------------------|-----------------|
| Hunan county boundary layer | Geospatial data set representing the county boundaries of Hunan | ESRI Shapefile |
| Hunan_2012.csv | Contains selected local development indicators for Hunan in 2012 | CSV |

## Installing and Launching the R Packages

The following R packages will be used in this exercise:

| **Package** | **Purpose** | **Use Case in Exercise** |
|-----------------|----------------------------|----------------------------|
| **sf** | Imports, manages, and processes vector-based geospatial data. | Handling vector geospatial data such as the Hunan county boundary layer in shapefile format. |
| **spdep** | Provides functions for spatial dependence analysis, including spatial weights and spatial autocorrelation. | Computing spatial weights and creating spatially lagged variables. |
| **tmap** | Creates static and interactive thematic maps using cartographic quality elements. | Visualizing regional development indicators and plotting maps showing spatial relationships and patterns. |
| **tidyverse** | A collection of packages for data science tasks such as data manipulation, visualization, and modeling. | Importing CSV files, wrangling data, and performing relational joins. |

To install and load these packages, use the following code:

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```


## Import Data and Preparation

In this section, we will perform 3 necessary steps to prepare the data for analysis.


### Import Geospatial Shapefile

In the code chunk below we will use [`st_read()`](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

### Import Aspatial csv File

Next, we will import *Hunan_2012.csv* into R by using `read_csv()` of **readr** package. The output is R dataframe class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### Perform Relational Join

he code chunk below will be used to update the attribute table of hunan’s SpatialPolygonsDataFrame with the attribute fields of hunan2012 dataframe. This is performed by using left_join() of dplyr package.


```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

### Visualizing Regional Development Indicator

To visualize the regional development indicator, we are going to prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 by using qtm() of tmap package.

```{r fig.width=16, fig.height=8}
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)

```

::: callout-note
**Observations**


The left plot uses equal interval classification to divide GDP per capita (GDPPC) into five evenly spaced ranges, ensuring each class spans the same value range. This method is ideal for continuous datasets like temperature or precipitation and is appreciated for its simplicity and ease of interpretation, especially for nontechnical audiences. However, it can be misleading for skewed datasets, as many data points may cluster in just one or two classes, leaving others sparsely populated or empty.

------------------------------------------------------------------------

In contrast, the right plot applies equal quantile classification, which divides the 88 counties of Hunan into five groups with roughly 17.6 counties each. This approach is useful for highlighting relative rankings, such as identifying the top 20% of counties by GDPPC. While it ensures balanced representation across classes, it can distort the data’s true distribution—counties with vastly different values may end up in the same class, or similar values may be split across different classes, exaggerating differences that aren’t statistically significant.

:::
