---
title: "In-class Exercise 2: Spatial Point Patterns Analysis"
author: "Shubham Sinha"
date: "2025-09-06"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
format: 
  html:
    code-fold: true
    code-summary: "code chunk"
    number-sections: true
    number-depth: 4
---

# Installing and loading R packages

## Install **maptools** package

**maptools** is retired and its binary is removed from CRAN. However, we can download it from [Posit Public Package Manager](https://packagemanager.posit.co/client/#/repos/cran/setup?snapshot=2023-10-13&distribution=windows&r_environment=ide) snapshots by using following code

```{r}
#| eval: false
install.packages("maptools",
                 repos = "https://packagemanager.posit.co/cran/2023-10-13")
```

> Include `#| eval: false` in the code chunk to avoid **maptools** being downloaded and installed repetitively every time the Quarto document is rendered.

## Install remaining packages

The code chunk below install and load following packages into R environment:

-   [**sf**](https://r-spatial.github.io/sf/), a relatively new R package specially designed to import, manage and process vector-based geospatial data in R.
-   [**spatstat**](https://spatstat.org/), which has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.
-   [**raster**](https://cran.r-project.org/web/packages/raster/) which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format.
-   [**maptools**](https://cran.r-project.org/web/packages/maptools/index.html) which provides a set of tools for manipulating geographic data. In this hands-on exercise, we mainly use it to convert *Spatial* objects into *ppp* format of **spatstat**.
-   [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) which provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API.

```{r}
pacman::p_load(sf, raster, spatstat, tmap, tidyverse)
```

# The data

Data sets in this exercise are as follows:

-   `CHILDCARE` (Point Feature Data)
    -   This dataset provides both location and attribute information of childcare centres across Singapore.
    -   Format: GeoJSON
    -   Source: [data.gov.sg](https://data.gov.sg/)
-   `MP14_SUBZONE_WEB_PL` (Polygon Feature Data)
    -   This dataset provide information of URA 2014 Master Plan Planning Subzone boundary data.
    -   Format: ESRI Shapefile.
    -   Source: [data.gov.sg](https://data.gov.sg/)
-   `CostalOutline` (Polygon Feature Data)
    -   This dataset shows the national boundary of Singapore.
    -   Format: ESRI Shapefile
    -   Source: [Singapore Land Authority](https://www.sla.gov.sg/)

We will use [st_read()](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package will be used to import these three geospatial data sets into R.

::: panel-tabset
## Childcare data

Since the `childcare_sf` simple feature data frame is in the WGS84 geodetic CRS, which is not ideal for geospatial analysis, the `st_transform()` function from the **sf** package is used to reproject the data to the SVY21 coordinate system during import.

```{r}
childcare_sf <- st_read("data/ChildCareServices.geojson") %>%
  st_transform(crs = 3414)
```

Let's verify the crs of the data frame to ensure we're using EPSG 3414.

```{r}
st_crs(childcare_sf)
```

## Master Plan 2014 data

Let's load the Master Plan Planning data using `st_read()` function

```{r}
mpsz_sf <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")
```

Let's check coordinate system of this data frame

```{r}
st_crs(mpsz_sf)
```

`mpsz_sf` is also using EPSG 9001 instead of 3414 which is suitable for CRS SVY21. Let's assign correct EPSG code using `st_set_crs()` then verify the output.

```{r}
mpsz_sf <- st_set_crs(mpsz_sf,3414)
st_crs(mpsz_sf)
```
:::

## working with `st_union()`

is used to derive the coastal outline sf tibble data.frame

```{r}
sg_sf <- mpsz_sf %>%
  st_union()
```

*sg_sf* will look similar to the figure below.

```{r}
plot(sg_sf)
```

# Creating ppp objects from sf data.frame

We can use `as.ppp()` of [**spatstat.geom**](https://cran.r-project.org/web/packages/spatstat.geom/spatstat.geom.pdf) package to derive an **ppp** object layer directly from a sf tibble data.frame.

```{r}
# Check geometry
print(st_geometry(childcare_sf))

# Check CRS
print(st_crs(childcare_sf))

# Convert manually if needed
childcare_ppp <- as.ppp(st_coordinates(childcare_sf), W = as.owin(sg_sf))
plot(childcare_ppp)
```

Let's reveal the properties of the newly created ppp objects using `summary()`.

```{r}
summary(childcare_ppp)
```

# Creating *owin* object from sf data.frame

We can use `as.owin()` of [**spatstat.geom**](https://cran.r-project.org/web/packages/spatstat.geom/spatstat.geom.pdf) package to create an **owin** object layer directly from a sf tibble data.frame.

```{r}
sg_owin <- as.owin(sg_sf)
plot(sg_owin)
```

Let's reveal the properties of the newly created owin objects using `summary()`.

```{r}
summary(sg_owin)
```

# Combining point events object and owin object

Now we'll create an ppp object by combining childcare_ppp and sg_owin then plot the output.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
plot(childcareSG_ppp)
```

# Kernel Density Estimation of Spatial Point Event

Before performing Kernel Density Estimation, we need to re-scale the unit of measurement from meter to kilometer.

```{r}
childcareSG_ppp.km <- rescale.ppp(childcareSG_ppp, 
                                  1000, 
                                  "km")

kde_childcareSG_adaptive <- adaptive.density(
  childcareSG_ppp.km, 
  method="kernel")
plot(kde_childcareSG_adaptive)
```

# Kernel Density Estimation

There are two ways to convert KDE output into grid object

::: panel-tabset
## maptools method

```{r}
library(spatstat.geom)
library(raster)
library(sp)

# Set background color
par(bg = '#E4D5C9')

# Convert spatstat image to raster
r <- raster(kde_childcareSG_adaptive)

# Convert raster to SpatialGridDataFrame
gridded_kde_childcareSG_ad <- as(r, "SpatialGridDataFrame")

# Plot using spplot
spplot(gridded_kde_childcareSG_ad)

```

## spatstat.geom method

```{r}
gridded_kde_childcareSG_ad <- as(
  kde_childcareSG_adaptive,
  "SpatialGridDataFrame")
spplot(gridded_kde_childcareSG_ad)
```
:::

## Visualising KDE using `tmap`

We can plot the output raster by using tmap functions.

```{r}
kde_childcareSG_ad_raster <- raster(gridded_kde_childcareSG_ad)
projection(kde_childcareSG_ad_raster) <- CRS("+init=EPSG:3414")
```

```{r}
tm_shape(kde_childcareSG_ad_raster) + 
  tm_raster(palette = "viridis") +
  tm_layout(legend.position = c("right", "bottom"), 
            frame = FALSE,
            bg.color = "#E4D5C9")
```

# Monte Carlo Simulation

In order to ensure reproducibility, it is important to include the code chunk below before using **spatstat** functions involve Monte Carlo simulation

```{r}
set.seed(2024)
```

# Reference

Kam, T. S. In-class Exercise 2: Spatial Point Patterns Analysis: spatstat methods. ISSS626 Geospatial Analytics and Applications.

-   <https://isss626-ay2025-26aug.netlify.app/lesson/lesson02/lesson02-sppa>
