---
title: "Hands-on Exercise 9A: Modelling Geographical Accessibility"
author: "Shubham Sinha"
date: "2025-11-06"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    number-depth: 4
---


## Overview

In this exercise, we will learn to model geographical accessibility using Hansen's potential model, Spatial Accessibility Measure (SAM), and other methods in R.

## Learning Outcome

-   Import GIS polygon data into R and save them as a simple feature data frame using the **sf** package.
-   Import aspatial data into R and save them as a simple feature data frame using the **sf** package.
-   Compute accessibility measures using Hansen's potential model and Spatial Accessibility Measure (SAM).
-   Visualize the accessibility measures using **tmap** and **ggplot2** packages.

## The Data

The following datasets will be used in this exercise:

| **Data Set** | **Description** | **Format** |
|----------------|-----------------------------------------|----------------|
| `MP14_SUBZONE_NO_SEA_PL` | URA Master Plan 2014 subzone boundary GIS data. | ESRI Shapefile |
| `hexagons` | A 250m radius hexagons GIS data created using the `st_make_grid()` function of the **sf** package. | ESRI Shapefile |
| `ELDERCARE` | GIS data showing the location of eldercare services, available in both ESRI shapefile and Google KML format. | ESRI Shapefile |
| `OD_Matrix` | A distance matrix with origin-destination information, including entry, network, and exit costs. | CSV |

> All the values of the cost related fields are in **metres**.

## Installing and Launching the R Packages

The following R packages will be used in this exercise:

| **Package** | **Purpose** | **Use Case in Exercise** |
|-----------------|---------------------------|----------------------------|
| **sf** | Handles spatial data; imports, manages, and processes vector-based geospatial data. | Importing and transforming geospatial datasets such as subzone boundaries, hexagons, and eldercare locations. |
| **SpatialAcc** | Provides functions for computing geographical accessibility measures. | Calculating Hansen's potential model, Spatial Accessibility Measure (SAM), and other accessibility metrics. |
| **tidyverse** | A collection of R packages for data science tasks like data manipulation, visualization, and modeling. | Wrangling and visualizing data, including importing CSV files and performing data transformations. |
| **tmap** | Creates static and interactive thematic maps using cartographic quality elements. | Visualizing accessibility measures on thematic maps. |
| **ggplot2** | Creates data visualizations using a layered grammar of graphics. | Visualizing statistical graphics such as histograms and boxplots of accessibility measures. |
| **ggstatsplot** | Enhances plots with statistical details and facilitates data visualization. | Creating statistically enriched plots for exploratory data analysis and comparing distributions. |
| **reshape2** | Provides tools to reshape data between wide and long formats. | Transforming data matrices into suitable formats for modeling. |

To install and load these packages, use the following code:

```{r}
pacman::p_load(tmap, SpatialAcc, sf, ggstatsplot, reshape2, tidyverse)
```

## Geospatial Data Wrangling

### Importing Geospatial Data

Three geospatial datasets—*MP14_SUBZONE_NO_SEA_PL*, *hexagons*, and *ELDERCARE*—are imported from the *data/geospatial* folder using the `st_read()` function from the **sf** package.

```{r echo=TRUE, eval=TRUE}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons")
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE")
```

### Updating CRS Information

We update the Coordinate Reference System (CRS) to EPSG:3414 for all datasets.

```{r, echo=TRUE, eval=TRUE}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
```

You can verify the CRS of *mpsz* using `st_crs()`:

```{r echo=TRUE, eval=TRUE}
st_crs(mpsz)
```

### Cleaning and Updating Attribute Fields

We remove redundant fields and add new ones. For *eldercare*, we add a `capacity` field, and for *hexagons*, we add a `demand` field, both with a constant value of 100.

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

::: callout-tip
For this exercise, a constant value of 100 is used. In practice, actual demand and capacity values should be used.
:::

## Aspatial Data Handling and Wrangling

### Importing Distance Matrix

We import the `OD_Matrix.csv` file using `read_csv()` from the **readr** package, which creates a tibble data frame called `ODMatrix`.

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

### Tidying the Distance Matrix

The imported *ODMatrix* organizes the distance matrix column-wise.

```{r}
head(ODMatrix)
```

However, most R modeling packages expect the matrix in a format where rows represent origins (from) and columns represent destinations (to).

We use `pivot_wider()` from the **tidyr** package to reshape the data from a long format to a wide format.

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  pivot_wider(names_from = destination_id, values_from = total_cost) %>%
  select(-origin_id)
```

### Converting Distance to Kilometers

Since the distances are in meters (due to the SVY21 projected coordinate system), we convert them to kilometers using the code below.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

## Modelling and Visualising Accessibility using Hansen Method

## Power Selection 2

The code chunk below uses `ac()` to compute the Hansen’s accessibility and saves the output in a data frame by using data.frame().

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km,
                            power = 2,
                            family = "Hansen"))
```

Let's have a quick review on the Hansen accessibility measure, after rename the column name to `accHansen`.

```{r}
colnames(acc_Hansen) <- "accHansen" 
```

We convert the data table into tibble format by using the code chunk below.

```{r}
acc_Hansen <- acc_Hansen %>%
  as_tibble()
```

```{r}
knitr::kable(head(acc_Hansen, 10), 
             align = 'c',  # column alignment: c=center, l=left, r=right
             format = "html")  # or "latex", "markdown", "pandoc"
```

Lastly, we join the `acc_Hansen` `tibble` with the `hexagon` `sf` data frame using `bind_cols()`. Because the `sf` object is on the **left** side of the operation, the resulting dataset automatically inherits the **sf** class, preserving its spatial attributes.

```{r}
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

```{r}
knitr::kable(head(hexagon_Hansen,10),
             align = "c",
            format = "markdown")
```


### Visualising Hansen’s Accessibility

#### Extracting Map Extend

First, we extract the extent of the hexagon sf layer using `st_bbox()` of sf package.

```{r}
mapex <- st_bbox(hexagons)
```

Next, we create a choropleth map with the **tmap** package to visualise eldercare service accessibility across Singapore.

```{r}
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

The map indicates that accessibility is highest in areas with clusters of eldercare centres, particularly in the Central Region and established heartland estates. In contrast, large peripheral areas exhibit zero accessibility.

### Statistical Graphic Visualisation

Next, we create a choropleth map with the **tmap** package to visualise eldercare service accessibility across Singapore.

In this section, we compare the distribution of Hansen accessiblity scores across URA Planning regions. To do so, we first append the main dataset `mpsz` to the `hexagon_Hansen` `sf` data frame using `st_join()`.

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz, join= st_intersects)
```

Then we use `ggplot()`, `geom_boxplot()` and `stat_summary()` , to create boxplots for each region and display the mean accessibility value.

```{r}
ggplot(data=hexagon_Hansen, 
       aes(y = log(accHansen), x= REGION_N)) +
  geom_boxplot() +
  stat_summary(fun="mean", colour ="red", size=0.5)
```

The boxplots of Hansen’s accessibility show that the Central Region has the highest average level of accessibility, followed by the North-East Region. In contrast, the West Region records the lowest average accessibility. The East Region displays the largest variability, as indicated by its long lower whiskers, suggesting uneven access to eldercare services across the region, possibly influenced by extensive non-residential areas, the airport.

The average Hansen accessibility scores rank the regions from highest to lowest as follows: Central, North-East, North, East, and West.


## Power Selection 0.5 

```{r}
acc_Hansen_05 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 0.5, 
                            family = "Hansen"))

colnames(acc_Hansen_05) <- "accHansen"
acc_Hansen_05 <- as_tibble(acc_Hansen_05)
hexagon_Hansen_05 <- bind_cols(hexagons, acc_Hansen_05)
```

```{r}
tm_shape(hexagon_Hansen_05,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```


```{r}
hexagon_Hansen_05 <- st_join(hexagon_Hansen_05, mpsz, join= st_intersects)
ggplot(data=hexagon_Hansen_05, 
       aes(y = log(accHansen), x= REGION_N)) +
  geom_boxplot() +
  stat_summary(fun="mean", colour ="red", size=0.5)
```
## Power Selection 1


```{r}
acc_Hansen_10 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 1.0, 
                            family = "Hansen"))

colnames(acc_Hansen_10) <- "accHansen"
acc_Hansen_10 <- as_tibble(acc_Hansen_10)
hexagon_Hansen_10 <- bind_cols(hexagons, acc_Hansen_10)
```

```{r}
tm_shape(hexagon_Hansen_10,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```


```{r}
hexagon_Hansen_10 <- st_join(hexagon_Hansen_10, mpsz, join= st_intersects)
ggplot(data=hexagon_Hansen_10, 
       aes(y = log(accHansen), x= REGION_N)) +
  geom_boxplot() +
  stat_summary(fun="mean", colour ="red", size=0.5)
```

## Power Selection 1.5

```{r}
acc_Hansen_15 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 1.5, 
                            family = "Hansen"))

colnames(acc_Hansen_15) <- "accHansen"
acc_Hansen_15 <- as_tibble(acc_Hansen_15)
hexagon_Hansen_15 <- bind_cols(hexagons, acc_Hansen_15)
```

```{r}
tm_shape(hexagon_Hansen_15,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```


```{r}
hexagon_Hansen_15 <- st_join(hexagon_Hansen_15, mpsz, join= st_intersects)
ggplot(data=hexagon_Hansen_15, 
       aes(y = log(accHansen), x= REGION_N)) +
  geom_boxplot() +
  stat_summary(fun="mean", colour ="red", size=0.5)
```

## Power Section 2.5


```{r}
acc_Hansen_25 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2.5, 
                            family = "Hansen"))

colnames(acc_Hansen_25) <- "accHansen"
acc_Hansen_25 <- as_tibble(acc_Hansen_25)
hexagon_Hansen_25 <- bind_cols(hexagons, acc_Hansen_25)
```

```{r}
tm_shape(hexagon_Hansen_25,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```


```{r}
hexagon_Hansen_25 <- st_join(hexagon_Hansen_25, mpsz, join= st_intersects)
ggplot(data=hexagon_Hansen_25, 
       aes(y = log(accHansen), x= REGION_N)) +
  geom_boxplot() +
  stat_summary(fun="mean", colour ="red", size=0.5)
```

:::callout-note
When varying the distance-decay power from 0.5 to 2.5 in the `ac()` function (Hansen accessibility), the overall magnitude of scores falls as the power increases. For example, with power = 0.5 the scores range from 0 to 1,158, whereas at power = 2.5 they range from 0 to 136. Intuitively, stronger distance-decay (higher power) penalises far-away opportunities more heavily, so aggregate accessibility declines. In areas where services are sparse or demand is strong relative to supply, accessibility is correspondingly lower.

:::

## OBSERVATIONS

Spatial Distribution of Eldercare Accessibility

Central and Eastern Regions: High Accessibility
Neighbourhoods such as Toa Payoh, Kallang, Queenstown, Bedok, and Tampines consistently register elevated accessibility scores across all power values. This pattern reflects a denser concentration of eldercare facilities and/or reduced travel distances within these zones.

Northern and Western Regions: Low Accessibility
In contrast, areas including Woodlands, Sembawang, Yishun, Jurong West, and Lim Chu Kang exhibit comparatively lower accessibility. This suggests either a sparse distribution of eldercare services or greater spatial separation between demand points and available capacity.

Effect of Power Value on Accessibility Sensitivity

- Low Power Values (e.g., 0.5): Accessibility is more uniformly distributed, as distant facilities retain substantial influence.

- Moderate Power Values (1 to 1.5): Spatial disparities become more pronounced, revealing clear distinctions between well-served and underserved regions.

- High Power Values (2 to 2.5): Accessibility becomes highly localized. Only areas in close proximity to eldercare facilities maintain high scores, with sharp declines observed as distance increases.


