---
title: "Take-Home Exercise 01"
author: "Shubham Sinha"
---

This is the TAKE HOME Ex01

## Loading the R Package

```{r}
pacman::p_load(sf,tmap,tidyverse,httr)
```

## Importing the ACRA Data

```{r} 
folder_path <- "data/aspatial/acra" 
file_list <- list.files(path=folder_path,
                        pattern ="^ACRA*.*\\.csv$",
                        full.names = TRUE)
acra_data <- file_list %>%
  map_dfr(read_csv)
```

## Saving ACRA data

```{r}
write_rds(acra_data,"data/rds/acra_data.rds")
```

## Tidying ACRA Data

write the code to do the following - 
1. Select target data using SSIC code
2. Derive the year and month fields from the registeration_incorporation_date field
3. tidy postal code values ensure avoid 5-digit postal code mistakes
4. select businesses registered in 2025

CODE HERE
bix_56111 <- acra_data %>%
select (1:24) %>%
filter (primary_ssic_code == 56111) %>%
rename (date=registeration_incorporation_date) %>%
mutate (date = as.Date(date),
YEAR = year(date),
MOTH_NUM = month(date),
MONTH_ABBR = month(date, label=TRUE, abbr = TRUE)) %>%
mutate(
postal_code = str_pad(postal_code,
width=6, side="left", pad="0")) %>%
filter(YEAR==2025)


## Geocoding

code in picture

run the code and will take really long time

## Tidying the geocoded data

## Appending the location information

biz_56111 = biz56111 %>%
left_join (found, 
by = c('postal_code'='POSTAL'))

## Saving the data

## Converting into SF data frame

biz_56111_sf <- st_as_sf(biz_56111, coords = c("X","Y"), crs=3414)


)






